---
title: "Two-stage Conditional Odds"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing a cutpoint and model structure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

> **Note.** This vignette was generated using AI under general human direction. At the time of generation, the contents have not been comprehensively reviewed by a human analyst.

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Fitting a Two-stage Conditional Odds model

Two-stage Conditional Odds (TsCO) models divide an ordinal response at a prespecified cutpoint, then allow the covariate effects below that cutpoint to differ from the effects governing the upper part of the outcome distribution. This vignette compares TsCO models against conventional proportional-odds (PO) and multinomial-regression (MR) models.

```{r load-packages}
library(tsco)
```

```{r load-data}
wine_env <- new.env(parent = emptyenv())
utils::data("wine", package = "ordinal", envir = wine_env)
wine <- as.data.frame(wine_env$wine)

# VGAM's multinomial model should receive an unordered outcome.
wine$rating_nominal <- factor(wine$rating, ordered = FALSE)
```

## Set the cutpoint from the study question

The `ordinal::wine` data come from a factorial experiment on wine bitterness.
Nine judges rated two bottles from each temperature-by-skin-contact condition,
producing 72 ratings. The ordered rating is a grouped version of a 0--100
bitterness score, with five intervals of width 20. The example is a
demonstration of model formulation rather than a definitive analysis. In
particular, the wine ratings are clustered by judge and bottle; the models below
treat the 72 ratings as conditionally independent and do not account for that
clustering.


```{r show-rating-scale}
data.frame(
  rating = levels(wine$rating),
  bitterness_score = c("0--<20", "20--<40", "40--<60", "60--<80", "80--100")
)
```

The TsCO approach assumes that the cutpoint is a priori selected. Information
criteria could be used to select a cutpoint in principle; we do not consider
that here. For illustration, we set `C = 4`: ratings 4 and 5 correspond to scores of at least 60 and define a high-bitterness range. The TsCO decomposition asks two distinct questions:

1. How do temperature and skin contact affect the chance of entering the
   high-bitterness range (ratings 4--5) rather than remaining below it?
2. Among wines below that range, how do they affect the relative probabilities
   of ratings 1, 2, and 3?

This is an illustrative target rather than a validated acceptability threshold.
If a real application has a clinical, regulatory, or sensory threshold, that
threshold—not a data-driven search over cutpoints—should determine `C`.

```{r tabulate-ratings}
with(wine, table(rating, temp, contact))
```

The table also identifies an important limitation: rating 5 is absent in both
cold treatment conditions. Fully multinomial fits therefore have sparse cells,
and estimates involving the highest rating warrant particular caution.

## Fit conventional ordinal models

The conventional PO model imposes one common effect of each predictor across
all cumulative logits. The conventional MR model instead allows a separate
predictor effect for each non-reference rating. They bracket the amount of
flexibility available without the scientifically motivated partition.

```{r fit-conventional-models}
fit_po <- ordinal::clm(rating ~ temp + contact, data = wine)

fit_mr <- VGAM::vglm(
  rating_nominal ~ temp + contact,
  family = VGAM::multinomial(),
  data = wine
)
```

## Fit TsCO models at the prespecified cutpoint

At `C = 4`, both component outcomes contain three categories. We can therefore
consider both a PO and an MR component in each stage. The four TsCO models
below vary the stage-1 model for ratings 1--3 and the stage-2 model for
`{rating < 4, rating 4, rating 5}`.

```{r fit-tsco-models}
fit_tsco_popo <- tsco(
  rating ~ temp + contact,
  data = wine,
  cutoff_level = "4",
  stage1 = "po",
  stage2 = "po"
)

fit_tsco_pomr <- tsco(
  rating ~ temp + contact,
  data = wine,
  cutoff_level = "4",
  stage1 = "po",
  stage2 = "multinomial"
)

fit_tsco_mrpo <- tsco(
  rating ~ temp + contact,
  data = wine,
  cutoff_level = "4",
  stage1 = "multinomial",
  stage2 = "po"
)

fit_tsco_mrmr <- tsco(
  rating ~ temp + contact,
  data = wine,
  cutoff_level = "4",
  stage1 = "multinomial",
  stage2 = "multinomial"
)
```

The ordinary MR and MR|C|MR specifications have the same number of free
parameters and describe the same model class under this factorization. Both
are retained below as a practical check. With sparse cells, iterative fitting
of equivalent parameterizations can yield small numerical differences; that is
not evidence that they have substantively different flexibility.

## Compare structures, not cutpoints

All models use the same 72 individual ratings, so their log likelihoods, AIC,
and BIC can be compared directly. The table uses the package's constant-free
individual-level likelihood for TsCO fits; for individual responses this is on
the same scale as the conventional-model likelihoods.

```{r compare-models}
model_fit <- function(model, label) {
  model_loglik <- stats::logLik(model)
  model_df <- attr(model_loglik, "df")

  if (is.null(model_df)) {
    model_df <- length(stats::coef(model))
  }

  data.frame(
    model = label,
    df = model_df,
    logLik = as.numeric(model_loglik)
  )
}

fit_comparison <- rbind(
  model_fit(fit_po, "Ordinary PO"),
  model_fit(fit_mr, "Ordinary MR"),
  model_fit(fit_tsco_popo, "PO | C = 4 | PO"),
  model_fit(fit_tsco_pomr, "PO | C = 4 | MR"),
  model_fit(fit_tsco_mrpo, "MR | C = 4 | PO"),
  model_fit(fit_tsco_mrmr, "MR | C = 4 | MR")
)

fit_comparison$AIC <- -2 * fit_comparison$logLik + 2 * fit_comparison$df
fit_comparison$BIC <- -2 * fit_comparison$logLik +
  log(nrow(wine)) * fit_comparison$df
fit_comparison <- fit_comparison[order(fit_comparison$BIC), ]
fit_comparison <- transform(
  fit_comparison,
  logLik = round(logLik, 2),
  AIC = round(AIC, 2),
  BIC = round(BIC, 2)
)
fit_comparison
```

For these data, BIC selects `r fit_comparison$model[1]`. The smallest TsCO
alternative is PO|C=4|PO, which gains some likelihood but adds two parameters.
Thus, these data do not provide enough support for the added two-stage
complexity under BIC.

The conventional PO and PO|C=4|PO models are **not generally nested**, so a likelihood-ratio test between them is invalid. Information criteria can compare these non-nested models because they are fitted to the same observed outcome likelihood.

By contrast, ordinary MR and MR|C|MR are alternative parameterizations of the
same multinomial model class. Multinomial logits are preserved by both
conditioning on the lower partition and collapsing that partition into one
category.

## Inspect the TsCO interpretation

Even when a conventional PO model is preferred by an information criterion,
the simplest TsCO model illustrates the stage-specific interpretation. Its
first stage models ratings 1--3 conditional on being below rating 4; its second
stage models the collapsed outcome consisting of below 4, rating 4, and rating
5.

```{r summarize-tsco-popo}
summary(fit_tsco_popo, joint_test = "none")
```

The following fitted probabilities make the two-stage result concrete for the
four treatment combinations.

```{r predict-treatment-combinations}
treatment_grid <- expand.grid(
  temp = levels(wine$temp),
  contact = levels(wine$contact),
  KEEP.OUT.ATTRS = FALSE
)

cbind(
  treatment_grid,
  round(predict(fit_tsco_popo, newdata = treatment_grid), 3)
)
```

## Practical workflow

In an applied analysis, first define `C` from the scientific or decision
context, check that enough observations occupy every relevant category, and
then compare conventional PO and MR models with TsCO structures at that one
cutpoint. Prefer the simplest adequately fitting model, while retaining a TsCO
model when stage-specific effects answer the substantive question and the data
support the additional parameters. Sparse outcome-by-covariate combinations,
as in this example, should limit how strongly model differences are
interpreted.

The prediction support check is marginal by factor level. With interaction
terms such as `x1 * x2`, it does not identify an absent combination when each
factor level is represented separately in the stage-1 data. Predictions for
such combinations should be treated as model-based extrapolations rather than
as empirically supported stage-1 predictions.

Complete or quasi-complete separation can also occur in either stage. A
likelihood-ratio test may be preferable to a Wald test when estimates are
unstable, but it does not make a finite maximum-likelihood estimate exist. Under
separation, Wald inference and the usual chi-squared calibration of a
likelihood-ratio statistic may both be unreliable; convergence diagnostics and
coefficient estimates should be examined before interpreting either test.

`summary()` prints a "Fit diagnostics" block when a stage fails to converge or
when a coefficient shows the extreme-estimate-and-extreme-standard-error
signature of separation. These data trigger it for the multinomial stages,
because rating 5 never occurs under cold temperature. That is also why the
MR|C|MR and ordinary MR log likelihoods above differ slightly: the likelihood
is nearly flat as the affected coefficients diverge, so the two
parameterizations stop at different points on the same plateau rather than
describing different models.
